### 2016-03-18 doc/draft.Rmd
methods draft

Madin-Darby canine kidney cells were provided by Arnold S. Monto (University of Michigan School of Public Health) and were maintained in Dulbecco's Modified Eagle Medium (DMEM, Invitrogen) with 10% fetal bovine serum (Gibco and HyClone), 25mM HEPES (Invitogen), and 0.1875% bovine serum albumin (Life Technologies) as in (MFE submitted). WSN33 was rescued from transfections using the 8 plasmid reverse genetic system containing the genomic segments for influenza A/WSN33 (pHW181-188) was a kind gift from Rober Webster (St. Jude's Children's Research Hospital) [@Hoffmann:2002ih]. A biological clone of influenza A/Puerto Rico/8/1934 was obtained from ATCC (VR-1469) and was cloned into the pHW2000 reverse genetic system as before [@Pauly:2015eq]

@@

Point mutants were made in the WSN33 background using the pHW2000 reverse genetics system (cite)  as previously described (MFE citation) . Briefly, we used  PCR mutagenesis to generate mutations at HA\_T1583G, HA\_G1006T, HA\_G542T, M\_T861G, M\_A541C, NA\_G1168T, NA\_C454T, NP\_A454C, NP\_A1160T, NS\_G227T, NS\_A809G, PA\_T964G, PA\_T237A, PA\_A1358T, PB1\_G599A, PB1\_G1764T, PB1\_T1288A, PB2\_A1854G, PB2\_A440T, PB2\_A1167T. Mutants were transfected and passaged once in MDCK cells prior to sequencing.

@@

Seven hundred and fifty nanograms of purified PCR products were sheared to an average size of 300-400 bp using a Covaris S220 focused ultrasonicator at the following settings : Intensity 4, Duty cycle 10%, Burst/second 200, time 80 seconds. Sequencing libraries were prepared from these fragmented products using the NEBNext Ultra DNA library prep kit (NEB E7370L) and NEBNext multiplex oligonucleotides for Illumina (NEB E7600S). We performed size selection during library preparation using Agencourt AMPure XP beads (Beckman Coulter A63881) with a target insert size of 300-400bp or 250 bp for sequencing on Illumina Miseq and Hiseq respectively.

@@

We extracted viral RNA from infected supernaturals using  QIAamp viral RNA minikits (Qiagen 5204) according to the manufactures' instructions, and generated cDNA using Superscript III one-step with HiFi platinum Taq (Invitrogen 12574). PCR products were purified using the Thermo Scientific PCR clean-up according to the manufactures' instructions.

#### PR8 WSN33 population 
WSN33 and PR8 isolates were plaque purified and passaged three times in MCDK cells. We varified the sequences of the viruses using Sanger sequencing.  cDNA was generated in segment specific reactions using : PB2-Forward-JT	(GCAGGTCAATTATATTCAATATGGAAA), PB2-Reverse-JT	(CAAGGTCGTTTTTAAACTATTCGACAC), PB1-Forward-JT (GCAGGCAAACCATTTGAATGG), PB1-Reverse-JT	(CAAGGCATTTTTTCATGAAGGACAAG), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG), NA-Forward-JT	(GCAGGAGTTTAAATGAATCCAAACC), NA-Reverse-JT	(CAAGGAGTTTTTTGAACAAACTACTTG), HA-Forward-JT	(GCAGGGGAAAATAAAAACAACCAAAAT), HA-Reverse-JT	(CAAGGGTGTTTTTCCTTATATTTCTGAA), NP-Forward-JT	(GCAGGGTAGATAATCACTCACAG), NP-Reverse-JT	(CAAGGGTATTTTTCTTTAATTGTCGTACT), M-Forward-JT	(GCAGGTAGATATTGAAAGATGAGTC), M-Reverse-JT	(CAAGGTAGTTTTTTACTCCAGCTCT), NS-Forward-JT	(GCAGGGTGACAAAGACATAATG), NS-Reverse-JT	(CAAAGGGTGTTTTTTATTATTAAATAAGCTG), which with the reaction conditions were adapted from [@Anonymous:A25Ka2tY]. Each 25 $\mu$ l reaction included 1 $\mu$ l of each primer (5 $\mu$ M) and 2 $\mu$ l of RNA template. Reaction conditions were 50$^\circ$ C (60 min), 94$^\circ$ C (2 min), followed by 30 cycles of 94$^\circ$ C (30 sec), 54$^\circ$ C (30 sec), and 68$^\circ$ C (3 min). PCR products were pooled to create equimolar mixes of WSN33 and PR8 segments. The WNS33 sample was then diluted into PR8 serially to yield mixtures in which WSN33 made up 5,2.5,1.25,and 0.63% of the population. We processed the samples for sequencing as below, and sequenced pooled libraries on an Illumina Miseq machine using 2X250 paired end reads.

To more closely mimic patient-derived viral populations we quantified the genome copy number of each of the 20 WSN33 point mutants (created above) and WSN33 wt using ... and the primers ... as described in ... . The mutants were mixed at equal concentrations and diluted to generate a mixture with each mutant present at 5% frequency and a total concentration of 10^5^ genomes per microliter. We diluted this population into the wt sample to create populations in which each mutant was present at 2, 1, 0.5,and 0.2% frequency all with a total concentration of 10^5^ genomes per micoliter. These 5 populations were diluted serially into basal media to generate samples with the same mutant frequencies as before but at nucleic acid concentrations of 10^4^ and 10^3^ genomes per microliter. Viral RNA was extracted from these samples as above and cDNA was generated in a multiplex RT-PCR as in [@Anonymous:A25Ka2tY] using the primers primers Uni12/Inf1 (GGGGGGAGCAAAAGCAGG), Uni12/Inf3 (GGGGGAGCGAAAGCAGG), and Uni13/Inf1 (5CGGGTTATTAGTAGAAACAAGG). Two multiplex RT-PCR reactions were run on the 10^5^ genomes per micoliter wt sample. We prepared libraries as before and used picogreen (.......) to quantify the concentration of each indexed library. _____ ng of each library was pooled together, gel isolated using __________, and sequenced using 2X125 paired end reads on an Illumina Hiseq machine.

### Plasmid controls

@@

Reads were aligned to either to PR8 (PR8 WSN33 data set) or WSN33 (patient derived conditions) using Bowtie2 (cite). Alignments were sorted and processed using Picard (......),variants were called using either DeepSNV or Lofreq and filtered using in house python and R scripts. We connected these steps into an analysis pipeline using bpipe. All commands needed to complete the analysis presented in this work are public ally available and free for download at .

>>>>

Madin-Darby canine kidney cells were provided by Arnold S. Monto (University of Michigan School of Public Health) and were maintained in Dulbecco's Modified Eagle Medium (DMEM, Invitrogen) with 10% fetal bovine serum (Gibco and HyClone), 25mM HEPES (Invitogen), and 0.1875% bovine serum albumin (Life Technologies) as in (MFE submitted). WSN33 was rescued from transfections using the 8 plasmid reverse genetic system containing the genomic segments for influenza A/WSN33 (pHW181-188), a kind gift from Rober Webster (St. Jude's Children's Research Hospital) [@Hoffmann:2002ih]. A biological clone of influenza A/Puerto Rico/8/1934 was obtained from ATCC (VR-1469) and was cloned into the pHW2000 reverse genetic system to be used as a sequencing control as before [@Pauly:2015eq].

@@

Point mutants were made in the WSN33 background using the pHW2000 reverse genetics system [@Hoffmann:2002ih] as previously described (MFE citation) . Briefly, we used  overlap PCR mutagenesis to generate mutations at HA\_T1583G, HA\_G1006T, HA\_G542T, M\_T861G, M\_A541C, NA\_G1168T, NA\_C454T, NP\_A454C, NP\_A1160T, NS\_G227T, NS\_A809G, PA\_T964G, PA\_T237A, PA\_A1358T, PB1\_G599A, PB1\_G1764T, PB1\_T1288A, PB2\_A1854G, PB2\_A440T, PB2\_A1167T.

@@

Seven hundred and fifty nanograms of purified PCR products were sheared to an average size of 300-400 bp using a Covaris S220 focused ultrasonicator using the following settings : Intensity:4, Duty cycle:10%, Burst/second: 200, Duration: 80 seconds. Sequencing libraries were prepared from these fragmented products using the NEBNext Ultra DNA library prep kit (NEB E7370L) and NEBNext multiplex oligonucleotides for Illumina (NEB E7600S). We performed size selection during library preparation using Agencourt AMPure XP beads (Beckman Coulter A63881) with a target insert size of 300-400bp or 250 bp for sequencing on Illumina Miseq and Hiseq respectively.

@@

We extracted viral RNA from infected supernaturants using  QIAamp viral RNA minikits (Qiagen 5204) according to the manufactures' instructions, and generated cDNA using Superscript III one-step with HiFi platinum Taq (Invitrogen 12574). PCR products were purified using the GeneJet PCR Purification Kit (ThermoFisher Scientific K0701) according to the manufactures' instructions.

#### PR8 WSN33 population

WSN33 and PR8 isolates were plaque purified and passaged three times in MCDK cells. We then varified the sequences of the these viruses using Sanger sequencing.  cDNA was generated in segment specific reactions using : PB2-Forward-JT	(GCAGGTCAATTATATTCAATATGGAAA), PB2-Reverse-JT	(CAAGGTCGTTTTTAAACTATTCGACAC), PB1-Forward-JT (GCAGGCAAACCATTTGAATGG), PB1-Reverse-JT	(CAAGGCATTTTTTCATGAAGGACAAG), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG), NA-Forward-JT	(GCAGGAGTTTAAATGAATCCAAACC), NA-Reverse-JT	(CAAGGAGTTTTTTGAACAAACTACTTG), HA-Forward-JT	(GCAGGGGAAAATAAAAACAACCAAAAT), HA-Reverse-JT	(CAAGGGTGTTTTTCCTTATATTTCTGAA), NP-Forward-JT	(GCAGGGTAGATAATCACTCACAG), NP-Reverse-JT	(CAAGGGTATTTTTCTTTAATTGTCGTACT), M-Forward-JT	(GCAGGTAGATATTGAAAGATGAGTC), M-Reverse-JT	(CAAGGTAGTTTTTTACTCCAGCTCT), NS-Forward-JT	(GCAGGGTGACAAAGACATAATG), NS-Reverse-JT	(CAAAGGGTGTTTTTTATTATTAAATAAGCTG), which with the reaction conditions were adapted from [@Anonymous:A25Ka2tY]. Each 25 $\mu$ l reaction included 1 $\mu$ l of each primer (5 $\mu$ M) and 2 $\mu$ l of RNA template. Reaction conditions were 50$^\circ\text{C}$ (60 min), 94$^\circ\text{C}$ (2 min), followed by 30 cycles of 94$^\circ\text{C}$ (30 sec), 54$^\circ\text{C}$ (30 sec), and 68$^\circ\text{C}$ (3 min). PCR products were pooled equimolarly to reconstituded cDNA genomes of WSN33 and PR8. WSN33 was then diluted into PR8 serially to yield mixtures in which WSN33 made up 5,2.5,1.25,and 0.63% of the population. We processed the samples for sequencing as above, pooled equal volumes of the indexed libraries, sequenced the pooled libraries on an Illumina Miseq machine using 2X250 paired end reads. A clonal plasmid control library was prepared from 8 plasmids containing PR8 genomic segments. These plasmids were mixed equimolarly and cDNA was generated using a multiplex RT-PCR with the primers Uni12/Inf1 (GGGGGGAGCAAAAGCAGG), Uni12/Inf3 (GGGGGAGCGAAAGCAGG), and Uni13/Inf1 (5CGGGTTATTAGTAGAAACAAGG) as in [@Anonymous:A25Ka2tY]. This library was then prepared in identical fashion to the experimental populations and was sequenced on the same Miseq lane.

#### Experimental intrahost population

We quantified the genome copy number of each of the 20 WSN33 point mutants (created above) and WSN33 wt using SuperScript III Platinum One-Step qRT-PCR (Invitrogen 11732). Primers and probes were designed by the by the CDC Influenza Branch for the universal detection of influenza A and B. The mutants were mixed at equal concentrations and diluted to generate a mixture with each mutant present at 5% frequency and a total concentration of 10^5^ genomes per microliter. We diluted this mixture into the wt sample to create populations in which each mutant was present at 2, 1, 0.5,and 0.2% frequency all with a total concentration of 10^5^ genomes per micoliter. These 5 populations were diluted serially into basal media to generate samples with the same mutant frequencies as before but at nucleic acid concentrations of 10^4^ and 10^3^ genomes per microliter. Viral RNA was extracted from these samples as above and cDNA was generated in a multiplex RT-PCR as in [@Anonymous:A25Ka2tY] using the primers Uni12/Inf1 (GGGGGGAGCAAAAGCAGG), Uni12/Inf3 (GGGGGAGCGAAAGCAGG), and Uni13/Inf1 (5CGGGTTATTAGTAGAAACAAGG). Two multiplex RT-PCR reactions were run on the 10^5^ genomes per micoliter wt sample. We prepared libraries as before and used  Quanti PicoGreen dsDNA quantification (ThermoFisher Scientific P7589) to quantify the concentration of each indexed library. Forty six ng of each indexed library was pooled together, gel isolated using the GeneJET Gel Extraction Kit (ThermoFisher Scientific K0691), and sequenced using 2X125 paired end reads on an Illumina Hiseq 2000 machine. A clonal control library was processed in an identical fashion starting from an equimolar mix of 8 plasmids containing the WSN33 genomic segments.

@@

Reads were aligned to either to PR8 (PR8 WSN33 data set) or WSN33 (patient derived conditions) using Bowtie2 [@Langmead:2012jh]. Alignments were sorted and processed using Picard (http://broadinstitute.github.io/picard/), variants were called using either DeepSNV [@Gerstung:1js] or Lofreq [@Wilm:2012br] and filtered using the Pysam module in python (https://github.com/pysam-developers/pysam) and R scripts. Bases with a Phred<30 were masked in the DeepSNV analysis. We connected these steps into an analysis pipeline using bpipe [@em:2012wd] which is available for download at .https://github.com/lauringlab/variant\_pipeline.  Additionally, all commands needed to complete the analysis presented in this work are publically available and free for download at https://github.com/lauringlab/Benchmarking\_paper . Finaly and interactive shiny app of our  benchmarking work can be 
downloaded at https://github.com/lauringlab/benchmarking\_shiny.
